---
title: "Reprex Tutorial"
author: "Julia Anstett"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## I've run into a problem with my R-Code and I don't know how to fix it. What Do I do?

If you need help getting unstuck, making a reprex, or reproducible example can help you share your code so you can ask for help. The goal of a reprex is to package your problematic code in such a way that other people can run it, then, hopefully, they can provide a solution.


## Why are Reprex's Helpful
Sometimes, going through the proccess of making a Reprex can help you solve your probelm before you even need to share your code.

80% of the time creating an excellent reprex reveals the source of your problem. It’s amazing how often the process of writing up a self-contained and minimal example allows you to answer your own question.

The other 20% of time you will have captured the essence of your problem in a way that is easy for others to play with. This substantially improves your chances of getting help if posting your issue online!


##  Installing the Reprex

```{r}

library(reprex)

```

## How To Create A Simple Reprex

There are two parts to creating a reprex:

First, you need to make your code reproducible. This means that you need to capture everything, i.e., include any library() calls and create all necessary objects. The easiest way to make sure you’ve done this is to use the reprex package.

Second, you need to make it minimal. Strip away everything that is not directly related to your problem. This usually involves creating a much smaller and simpler R object than the one you’re facing in real life or even using built-in data.


# Step 1: Write a piece of code, make sure you include all libraries that you need to execute your code. Make sure you load your reprex package first.

```{r, message=FALSE, error=TRUE}

library (reprex)
library (tidyverse)

x <- 1
y <- 2

z

```

# Step 2: copy the code for Reprex into your clipboard

# Step 3: enter reprex() into your console

```{r, eval=FALSE}

reprex()

```

Once you run reprex(), your code, its outputs, and any error messages generated will be saved to your clipboard. Note that this will be converted to R-Markdown format.

## Posting A Reprex
Then you can paste your code to Github, Stackoverflow, or any website that can interpret R-Markdown code.

## Tips on writing a good reprex

- Explicitly list all packages required to run your code.
- Make sure all objects needed to run your code are made available.


## Additonal resources
- https://www.tidyverse.org/help/
- https://reprex.tidyverse.org/
- https://reprex.tidyverse.org/articles/reprex-dos-and-donts.html
- https://www.youtube.com/watch?v=MmTPhGQWPUo&t=2s

