---
title: "Intermediate R and RStudio"
author: "Cathy Yan"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: [DESCRIPTION]
---

```{r setup, include = FALSE}
# General learnr setup
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
library(educer)
# Helper function to set path to images to "/images" etc.
setup_resources()

# Tutorial specific setup
library(dplyr)
library(readr)
library(tidyverse)
```

## Learning objectives

Here's what you'll learn from each section of this tutorial:

Loops in R:

- Explain the mechanisms of for and while loops
- Implement for and while loops
- Recognize and understand how nested loops work

Conditionals in R:

- Write and evaluate and/or statements
- Identify the components of if/else statements
- Implement conditionals within loops

Functions in R:

- Write functions that operate on arguments
- Understand the role of helper functions
- Implement helper functions

Manipulating Data Frames:

- Implement the pipe operator
- Use the `select()`, `filter()`, and `mutate()` functions appropriately
- Operate on individual columns within a data frame
- Change the data type of columns in a data frame
- Replace NAs in a vector or data frame

Other ideas: Tidying data, joining data frames, loading in data from an external source

## Review

Before proceeding, complete the five questions below reviewing content from Introduction to R and RStudio Fundamentals. If any of the concepts seem unfamiliar, please revisit the previous tutorial.

```{r using-libraries, echo=FALSE}
quiz(question("In no particular order, which actions are required before using functions from a new package?",
      answer("Install the package", correct=TRUE),
      answer("Download vignettes"),
      answer("Load the package", correct=TRUE),
      answer("Install dependencies"))
)
```

Pull up the help page for the function `quantile()`.

```{r q-help, exercise=TRUE}
# your code here
```

```{r q-help-solution}
?quantile
```

```{r review, echo=FALSE}
quiz(
  question("What types of arguments can be passed to `quantile()`?",
    answer("Logical", correct=TRUE),
    answer("Numeric", correct=TRUE),
    answer("Logical Vector"),
    answer("Numeric Vector", correct=TRUE),
    answer("Text"))
)
```

Create a vector with the numbers 9, 27, 15, 74, 36, 4, and 49 and assign it to the variable `x`. Calculate the quantiles for `x`.

```{r q-vector, exercise=TRUE}
# your code here
```

```{r q-vector-hint-1}
x <- c(9, 27, 15, 74, 36, 4, 49)
```

```{r q-vector-solution}
x <- c(9, 27, 15, 74, 36, 4, 49)
quantile(x)
```

```{r q-type1, echo=FALSE}
quiz(
  question("What is the 75% quantile for `x`?",
    answer("36"),
    answer("50.25"),
    answer("42.5", correct=TRUE))
)
```

```{r q-type7, echo=FALSE}
quiz(
  question("what is the 25% quantile for `x` if argument type = 1? ",
    answer("27"),
    answer("12"),
    answer("15"),
    answer("9", correct = TRUE))
)
```
```{r q-25, echo=FALSE}
quiz(
  question("what is the 25% quantile for `x` if argument type = 7? ",
    answer("27"),
    answer("9"),
    answer("15"),
    answer("12", correct = TRUE))
)

```




