---
title: "Intermediate R and RStudio"
author: "Cathy Yan"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: [DESCRIPTION]
---

```{r setup, include = FALSE}
# General learnr setup
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
library(educer)
# Helper function to set path to images to "/images" etc.
setup_resources()

# Tutorial specific setup
library(dplyr)
library(readr)
library(tidyverse)
```

## Learning objectives

Here's what you'll learn from each section of this tutorial:

Loops in R:

- Explain the mechanisms of for and while loops
- Implement for and while loops
- Recognize and understand how nested loops work

Conditionals in R:

- Write and evaluate and/or statements
- Identify the components of if/else statements
- Implement conditionals within loops

Functions in R:

- Write functions that operate on arguments
- Understand the role of helper functions
- Implement helper functions

Manipulating Data Frames:

- Implement the pipe operator
- Use the `select()`, `filter()`, and `mutate()` functions appropriately
- Operate on individual columns within a data frame
- Change the data type of columns in a data frame
- Replace NAs in a vector or data frame

Other ideas: Tidying data, joining data frames, loading in data from an external source

## Review

Before proceeding, complete the five questions below reviewing content from Introduction to R and RStudio Fundamentals. If any of the concepts seem unfamiliar, please revisit the previous tutorial.

```{r using-libraries, echo=FALSE}
quiz(question("In no particular order, which actions are required before using functions from a new package?",
      answer("Install the package", correct=TRUE),
      answer("Download vignettes"),
      answer("Load the package", correct=TRUE),
      answer("Install dependencies"))
)
```

Pull up the help page for the function `quantile()`.

```{r q-help, exercise=TRUE}
# your code here
```

```{r q-help-solution}
?quantile
```

```{r review, echo=FALSE}
quiz(
  question("What types of arguments can be passed to `quantile()`?",
    answer("Logical", correct=TRUE),
    answer("Numeric", correct=TRUE),
    answer("Logical Vector"),
    answer("Numeric Vector", correct=TRUE),
    answer("Text"))
)
```

Create a vector with the numbers 9, 27, 15, 74, 36, 4, and 49 and assign it to the variable `x`. Calculate the quantiles for `x`.

```{r q-vector, exercise=TRUE}
# your code here
```

```{r q-vector-hint-1}
x <- c(9, 27, 15, 74, 36, 4, 49)
```

```{r q-vector-solution}
x <- c(9, 27, 15, 74, 36, 4, 49)
quantile(x)
```

```{r q-type1, echo=FALSE}
quiz(
  question("What is the 75% quantile for `x`?",
    answer("36"),
    answer("50.25"),
    answer("42.5", correct=TRUE))
)
```

```{r q-type7, echo=FALSE}
quiz(
  question("what is the 25% quantile for `x` if argument type = 1? ",
    answer("27"),
    answer("12"),
    answer("15"),
    answer("9", correct = TRUE))
)
```
```{r q-25, echo=FALSE}
quiz(
  question("what is the 25% quantile for `x` if argument type = 7? ",
    answer("27"),
    answer("9"),
    answer("15"),
    answer("12", correct = TRUE))
)

```

## Writing Function in R

In the previous tutorial, you learned how to use preexisting functions. Now, you will write your own.

### Why Write Functions?

When you're performing the same set of transformations on the same kinds of inputs, you should consider writing a function to automate the task. Functions make your code:

- Reproducible
- Concise
- Easier to understand

### Syntax for Writing Functions

Functions take on the following form:

```{r fun-ex, eval=FALSE, error=TRUE}
myfunction <- function(arg1, arg2){
  myoutput <- myaction(arg1, arg2)
  return(myoutput)
}
```

Let's break down each component while writing a function that divides numbers by two:

#### Naming your function

In the previous tutorial, you used the function `sum()` to add a vector of numbers. You used the word "sum", the function's name, to call for it. Similarly, your own function also needs a name to be used. In our example above, the function's name is "myfunction", and you would call for it by typing `myfunction()`.

#### Using the `function()` function

Defining a function uses the same syntax as defining a variable, so how does R tell them apart? Using the `function()` function identifies your function name as such rather than as a variable.

To begin writing our function, which we'll name `halve()`, fill in the blanks below:

```{r fun-start, exercise = TRUE, eval=FALSE, error=TRUE}
... <- ...(){}
```

```{r fun-start-solution}
halve <- function(){}
```

#### Setting arguments

In the review quiz, you passed a vector of numbers as an argument to the function `quantile()`. Similarly, you'll need to be able to pass arguments to you own function. To do this, you'll need to name each of your arguments, and list them within the `function()` function. In the example above, we have two arguments: `arg1` and `arg2`. 

Right now, in order for the function to work, the user will need to provide both arguments. However, you can also set default values for arguments. The syntax is as follows:

```{r fun-ex2, eval=FALSE, error=TRUE}
myfunction <- function(arg1, arg2 = 5){
  myoutput <- myaction(arg1, arg2)
  return(myoutput)
}
```

Now, the user is only required to provide `arg1`. If they don't provide `arg2`, the function will run using the default value of 5. If `arg2` is provided, the input value overrides the default (i.e. if the user specifies a value of 7 for `arg2`, the function will use that instead of 5, the default).

For our function, we'll only need one argument, which we'll call `num`. It doesn't make sense for it to have a default value since there's no universal number people want to halve, so we won't include one.

```{r fun-arg, exercise = TRUE, eval = FALSE, error = TRUE}
halve <- function(...){}
```

```{r fun-arg-solution}
halve <- function(num){}
```

#### Operating on arguments

The function's body is enclosed by curly braces. Within, the arguments are transformed through one or more operations. In this case, there is only one operation, `myaction()`, which acts upon both arguments. The output of `myaction()` is assigned to the variable `myoutput`. This variable assignment works in the same way as usual, except for one caveat: `myoutput` cannot be used outside of the function. You will not see it show up in your environment.

We want our function to halve, or divide by two, any number we provide. We'll name our output variable `half`.

```{r fun-act, exercise = TRUE, eval = FALSE, error = TRUE}
halve <- function(num){
  half <- ...
}
```

```{r fun-act-solution}
halve <- function(num){
  half <- num / 2
}
```

#### Returning outputs

When you define a variable normally, you need to call on the variable to see its value. Within functions, the sample principle applies. Although `myoutput` is assigned a value, `myfunction()` will not actually return it unless you explicitly tell it to do so. In this example, the `return()` function is used to accomplish this. Let's return `half`:

```{r fun-ret, exercise = TRUE, eval = FALSE, error = TRUE}
halve <- function(num){
  half <- num / 2
  ...
}
```

```{r fun-ret-solution}
halve <- function(num){
  half <- num / 2
  return(half)
}
```

### Helper Functions

Of course, all of the functions you'll need will be much more complex than `halve()`. In fact, some will be so complex that you'll need *helper functions* to complete the task. In the example below, we have a function that adds 5 to the average of a list of numbers:

```{r num-list, exercise = TRUE}
avg_minus_five <- function(num_list){
  output <- mean(num_list) + 5
  return(output)
}

list = c(1, 2, 3, 4, 5, 6)

answer <- avg_minus_five(list)
answer
```

Did you notice anything different with `avg_minus_five()` compared to `halve()`? In the former, we're using another function, `mean()`, inside of the function we wrote. Instead of writing code to calculate the average from scratch, we're using `mean()` to help us get to our desired output in a quicker and more concise way. In this case, `mean()` is our helper function.

### This section may be more appropriate later?

Let's walk through the process for writing and implementing a helper function. Below, we have a long function. Can you guess what it does?

```{r long-func, exercise = TRUE}

```

### Check Your Understanding



## Manipulating Data Frames

### Transforming Data

The functions you have implemented so far all perform operations on one piece of data at a time. For example, `halve()` takes a single number and divides it by two, but can it be used on columns in a data frame?

```{r df, include=FALSE}
grades <- data.frame(
  name = c("John", "Jane", "Charles", "Amy", "Joe"),
  score = c(30, 80, 65, 92, 73),
  passing = c(FALSE, TRUE, TRUE, TRUE, TRUE)
)
```

We're going to be using our dataset with students' grades from the last tutorial:

```{r df2, exercise=TRUE, exercise.setup="df"}
grades
```

Try to divide all of the students' grades by half using the `halve()` function. The `$` symbol is used to select columns from a data frame.

```{r halve-col, exercise = TRUE, error=TRUE, exercise.setup="df"}
halve <- function(num){
  half <- num / 2
  return(half)
}

halve(grades$score)
```





### Replacing NAs

In the last tutorial, you were introduced to NAs, which denote missing data.

#### Removing rows with NAs

The simplest way to deal with NAs is to remove them. There are two options for this route: 

- Removing all rows with any NAs
- Removing rows with NAs in specific columns

Removing NAs is quick and easy, but real life data can have a large proportion of NAs, leading to significant data loss. 

#### Replacing NAs with a fixed value

- Replacing with the mean
- Replacing with 0/most common value

#### Replacing NAs based on neighbours

- Replacing with last occurring value
- Replacing with mean of two neighbouring values









