---
title: "Intermediate R and RStudio"
author: "Cathy Yan"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: [DESCRIPTION]
---

```{r setup, include = FALSE}
# General learnr setup
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
library(educer)
# Helper function to set path to images to "/images" etc.
setup_resources()

# Tutorial specific setup
library(dplyr)
library(readr)
library(tidyverse)
```

## Learning objectives

Here's what you'll learn from each section of this tutorial:

Loops in R:

- Explain the mechanisms of for and while loops
- Implement for and while loops
- Recognize and understand how nested loops work

Conditionals in R:

- Write and evaluate and/or statements
- Identify the components of if/else statements
- Implement conditionals within loops

Functions in R:

- Write functions that operate on arguments
- Understand the role of helper functions
- Implement helper functions

Manipulating Data Frames:

- Implement the pipe operator
- Use the `select()`, `filter()`, and `mutate()` functions appropriately
- Operate on individual columns within a data frame
- Change the data type of columns in a data frame
- Replace NAs in a vector or data frame

Other ideas: Tidying data, joining data frames, loading in data from an external source

## Review

Before proceeding, complete the five questions below reviewing content from Introduction to R and RStudio Fundamentals. If any of the concepts seem unfamiliar, please revisit the previous tutorial.

```{r using-libraries, echo=FALSE}
quiz(question("In no particular order, which actions are required before using functions from a new package?",
      answer("Install the package", correct=TRUE),
      answer("Download vignettes"),
      answer("Load the package", correct=TRUE),
      answer("Install dependencies"))
)
```

Pull up the help page for the function `quantile()`.

```{r q-help, exercise=TRUE}
# your code here
```

```{r q-help-solution}
?quantile
```

```{r review, echo=FALSE}
quiz(
  question("What types of arguments can be passed to `quantile()`?",
    answer("Logical", correct=TRUE),
    answer("Numeric", correct=TRUE),
    answer("Logical Vector"),
    answer("Numeric Vector", correct=TRUE),
    answer("Text"))
)
```

Create a vector with the numbers 9, 27, 15, 74, 36, 4, and 49 and assign it to the variable `x`. Calculate the quantiles for `x`.

```{r q-vector, exercise=TRUE}
# your code here
```

```{r q-vector-hint-1}
x <- c(9, 27, 15, 74, 36, 4, 49)
```

```{r q-vector-solution}
x <- c(9, 27, 15, 74, 36, 4, 49)
quantile(x)
```

```{r q-type1, echo=FALSE}
quiz(
  question("What is the 75% quantile for `x`?",
    answer("36"),
    answer("50.25"),
    answer("42.5", correct=TRUE))
)
```

```{r q-type7, echo=FALSE}
quiz(
  question("what is the 25% quantile for `x` if argument type = 1? ",
    answer("27"),
    answer("12"),
    answer("15"),
    answer("9", correct = TRUE))
)
```
```{r q-25, echo=FALSE}
quiz(
  question("what is the 25% quantile for `x` if argument type = 7? ",
    answer("27"),
    answer("9"),
    answer("15"),
    answer("12", correct = TRUE))
)

```

## Writing Function in R

In the previous tutorial, you learned how to use preexisting functions. Now, you will write your own.

### Why Write Functions?

When you're performing the same set of transformations on the same kinds of inputs, you should consider writing a function to automate the task. Functions make your code:

- Reproducible
- Concise
- Easier to understand

### Syntax for Writing Functions

Functions take on the following form:

```{r fun-ex, eval=FALSE, error=TRUE}
myfunction <- function(arg1, arg2){
  myoutput <- myaction(arg1, arg2)
  return(myoutput)
}
```

Let's break down each component:

#### Naming your function

In the previous tutorial, you used the function `sum()` to add a vector of numbers. You used the word "sum", the function's name, to call for it. Similarly, your own function also needs a name to be used. In our example above, the function's name is "myfunction", and you would call for it by typing `myfunction()`.

#### Using the `function()` function

Defining a function uses the same syntax as defining a variable, so how does R tell them apart? Using the `function()` function identifies your function name as such rather than as a variable. 

#### Setting arguments

In the review quiz, you passed a vector of numbers as an argument to the function `quantile()`. Similarly, you'll need to be able to pass arguments to you own function. To do this, you'll need to name each of your arguments, and list them within the `function()` function. In the example above, we have two arguments: `arg1` and `arg2`. 

Right now, in order for the function to work, the user will need to provide both arguments. However, you can also set default values for arguments. The syntax is as follows:

```{r fun-ex, eval=FALSE, error=TRUE}
myfunction <- function(arg1, arg2 = 5){
  myoutput <- myaction(arg1, arg2)
  return(myoutput)
}
```

Now, the user is only required to provide `arg1`. If they don't provide `arg2`, the function will run using the default value of 5. If `arg2` is provided, the input value overrides the default (i.e. if the user specifies a value of 7 for `arg2`, the function will use that instead of 5, the default).

#### Operating on arguments

The function's body is enclosed by curly braces. Within, the arguments are transformed through one or more operations. In this case, there is only one operation, `myaction()`, which acts upon both arguments. The output of `myaction()` is assigned to the variable `myoutput`. This variable assignment works in the same way as usual, except for one caveat: `myoutput` cannot be used outside of the function. You will not see it show up in your environment.  

#### Returning outputs

When you define a variable normally, you need to call on the variable to see its value. Within functions, the sample principle applies. Although `myoutput` is assigned a value, `myfunction()` will not actually return it unless you explicitly tell it to do so. In this example, the `return()` function is used to accomplish this.

### Helper Functions

Need green light on loops first.

### Check Your Understanding

Need green light on loops first.

## Manipulating Data Frames

### Replacing NAs












