The data used throughout this module were collected as part of an on-going oceanographic time series program in Saanich Inlet, a seasonally anoxic fjord on the East coast of Vancouver Island, British Columbia.

The data that you will use in R are 16S amplicon profiles of microbial communities at several depths in Saanich Inlet from one time point in this series (August 2012). These ~300 bp sequences were processed using [mothur](https://www.mothur.org/wiki/Main_Page) to yield 97% (approximately species-level) operational taxonomic units (OTUs). 

`combined` is a comma-delimited table of counts of four OTUs in each sample, normalized to 100,000 sequences per sample and the corresponding conditions of each sample (Depth, NO2, NO3 etc). 

For a brief introduction to these data, see Hallam SJ et al. 2017. Monitoring microbial responses to ocean deoxygenation in a model oxygen minimum zone. Sci Data 4: 170158 [doi:10.1038/sdata.2017.158](https://www.nature.com/articles/sdata2017158).

Click the button below to save a copy of this data set to your computer:

```{r echo = FALSE}
# Download button shiny app UI
fluidRow(
  column(12, align = "center", downloadButton("downloadData", "Download"))
)

```

```{r context = "server"}
# Download button shiny app server
output$downloadData <- downloadHandler(
  filename = "combined.csv",
  content = function(file) {
    write_csv(combined, file)
  }
)
```

