---
title: "tidyverse advanced"
author: "michelle kang"
date: "07/02/2020"
output: learnr::tutorial
runtime: shiny_prerendered
description: this file contains the third of three data wrangling tutorials using the tidyverse package in R. This tutorial introduces joining datatables using the inner_join function and converting data from long format to wide format using the pivot functions. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
library(tidyverse)
```

## Introduction
Let's first load the tidyverse package into our environment

**R v3.4 or newer**
```{r message=FALSE}
library(tidyverse)
library(educer)
```

After completing this tutorial you will be able to:

-use the left_join function to combine two data tables

## Refresher 

From the previous 2 tutorials you should be comfortable using the:

- `select(dataframe, specified columns separated by commas) ` : select to work with only specified columns from our data table 

- `filter(dataframe, boolean condition)` : select specific rows in dataframe based on a logical condition of a column value.

- `slice(dataframe, specified rows separated by commas OR range of rows separated by ":")` : select a subset of observations (rows) by their position in the dataframe

- the `rename(dataframe, new_column_name = old_column_name)` function to assign more intuitive names for columns

- the `mutate(dataframe, new_column = transformation of old_column)` function to apply a transformation to values in one column and assign new values to a new column

- piping with %>% to chain commands together

- the `summarise` function to calculate summary statistics such as mean and standard deviation for groups of data

```{r echo=FALSE, results=FALSE}
pdat <- geochemicals
temp <- pdat %>% 
  select(Depth, CTD_O2) %>% 
  filter(Depth == 100) %>% 
  rename(Oxygen = CTD_O2) %>% 
  mutate(Oxygen = Oxygen*32) %>%
  filter(!is.na(Oxygen))
```

Exercise:

We will be using the "geochemicals" dataset included in the educer package. This dataframe contains time series observations on the water column chemistry. Learn more about the geochemicals dataset by running "?educer" in your R console. The geochemicals dataset has been loaded into "dat" for you as a start.

1. Select the Depth and CTD_O2 variables from geochemicals dataset
2. Filter to 100 m samples
3. Rename the O2 variable to `Oxygen` using rename
4. Transform Oxygen from micromoles/L to micrograms/L using mutate (multiply Oxygen by 32)
    - You should now have a [`r dim(temp)`] data frame.
5. Calculate the mean and standard deviation of Oxygen in micrograms/L for this subset of data.

This should result in `r mean(temp$Oxygen)` $\pm$ `r sd(temp$Oxygen)`

```{r refresher_exercise, exercise = TRUE, exercise.lines = 5}
dat <- geochemicals

```

Write a one line command that filters the geochemicals dataset for all geochemical mean values (*hint* use the "starts_with" helper function)

```{r refresher_startswith, exercise = TRUE, exercise.lines = 5}
```

## Join Tables 

Dyplr has built in functions to quickly join two tables. In this tutorial we will be focusing on `inner_join`. `inner_join` takes in two tables, x and y, and return all rows from x where there are matching values in y, and all columns from x and y. 

If there are multiple matches between x and y, *all combination of the matches are returned*.

for example, a table x:
```{R echo=FALSE, warning=FALSE, message=FALSE}
metadata <- read_tsv("Saanich_metadata.txt", col_names = TRUE)
head(metadata)
```

and a table y:
```{r echo=FALSE, warning=FALSE, message=FALSE}
otu_data <- read_tsv("Saanich_OTU.txt", col_names = TRUE)
otu_data <- rename(otu_data, sample = X1)
head(otu_data)
```

can be joined by their mutual column "sample" like this:
```{r warning=FALSE, message=FALSE}
metadata <- read_tsv("Saanich_metadata.txt", col_names = TRUE)
otu_data <- read_tsv("Saanich_OTU.txt", col_names = TRUE)
otu_data <- rename(otu_data, sample = X1)
otu_metadata <- inner_join(metadata, otu_data, by = "sample")
otu_metadata
```

as you may have noticed, this is how the "Saanich_OTU_Metadata.csv" file from the introductory tutorial was generated. 

we will select the first 15 columns to work with fewer OTUs:
```{R}
subset_otu_metadata <- select(otu_metadata, 1:15)
```

perform an inner join on the following two data frames of fruits students brought for lunch in classroom A and classroom B.
```{r join_table_exercise, exercise = TRUE, exercise.lines = 10}
fruits = c("apple", "orange", "pear", "grape", "orange")
student_A = c("Joe", "Maria", "Tom", "Sally", "James")
classroom_A <- data.frame(fruits,student_A)

fruits = c("apple", "orange", "grape", "pear", "orange")
student_B = c("Jason", "Mary", "Tory", "Sarah", "Josh")
classroom_b <- data.frame(fruits,student_B)


```

```{r table_join_quiz, echo = FALSE}
quiz(
  question("How many rows does the resulting dataframe have?",
           answer("5"),
           answer("6"),
           answer("7", correct = TRUE)
           ),
  question("What happens if you decrease the size of the classroom A dataset by deleting one orange and one student?",
           answer("ERROR both data frames have to be equal in size for them to be joined"),
           answer("The resulting joint table will have 5 rows", correct = TRUE),
           answer("The resulting joint table will have 6 rows")
           )
)
```

