---
title: "Intro to R quiz"
author: "Andrew Wilson"
date: "20/01/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Which of the following is a good variable name?
  * x
  * 1st_pass
  * sum
  * _raw_saanich_otus_

2. What is the name of the pane in RStudio where you should write most of your code?
  * Console
  * _Script_
  * Environment
  * Files

3. Which `dplyr` function can you use to select rows based on values in a column?
  * select
  * slice
  * mutate
  * _filter_

4. Which `col_names` argument in read_csv will break?
  * c("column_1", "column_2", "column_3", "column_4")
  * TRUE
  * FALSE
  * _c(1, 2, 4, 5)_
  
5. Adding which line to the beginning of the script would fix this error?
```{r, eval = TRUE, warning=TRUE, error=TRUE}
saanich_abund <- read_csv("../../../MICB301/data/Saanich_abund.txt")
```
  * _library(tidyverse)_
  * dim(saanich_abund)
  * setwd("~")
  * colnames(saanich_abund)

6. What are the main benefits of R projects?
  * Creates a virtual environment for R
  * Collects all of your data, outputs, and scripts in the same working directory
  * They are units of shareable code, containing functions that enhance analysis
  * They are like applications for your phone
